---
title: "Quiz 1"
author: "George Papadopoulos </br> pgeorgios8@gmail.com"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---
<style>body{text-align: justify}</style>

## Question 1

Consider influenza epidemics for two parent heterosexual families. Suppose that the probability is 17% that at least 
one of the parents has contracted the disease. The probability that the father has contracted influenza is 12% while
the probability that both the mother and father have contracted the disease is 6%. What is the probability that the
mother has contracted influenza?

<u>answer</u>

Let A be the probability of the father being infected by influenza and B the mother's probability of having 
contracted the virus.
\[
\begin{eqnarray*}
\mathbb{P}\left(A\bigcup B\right) &=& \mathbb{P}\left(A\right) + \mathbb{P}\left(B\right) - 
\mathbb{P}\left(A\bigcap B\right)\\
0.17 &=& 0.12 + \mathbb{P}\left(B\right) - 0.06 \\
\mathbb{P}\left(B\right) &=& 0.11
\end{eqnarray*}\]

## Question 2

A random variable, $\mathbb{\textrm{X}}$$ is uniform, a box from 0 to 1 of height 1. 
(So that its density is $f(x) = 1)$ for $0\leq x \leq 1$. What is its 75th percentile?

<u>answer</u>

```{r a2,comment=''}
qunif(0.75)
```

## Question 3

You are playing a game with a friend where you flip a coin and if it comes up heads you give her x dollars and if it
comes up tails she gives you y dollars. The probability that the coin is heads is p (some number between 0 and
1). What has to be true about x and y to make so that both of your expected total earnings is 0 The game would then be
called “fair”.

<u>answer</u>

\[\mathbb{E}\left[\text{earnings}\right] = -p\cdot X + (1-p)\cdot Y = 0\Rightarrow\frac{p}{1-p} = \frac{y}{x}\]

## Question 4

A density that looks like a normal density (but may or may not be exactly normal) is exactly symmetric about zero.
(Symmetric means if you flip it around zero it looks the same.) What is its median? 

<u>answer</u>

$\color{red}0$, 50% of it must fall below 0 and the other 50% above it.

## Question 5

Consider the following PMF shown below in R

```{r q5,comment=''}
x <- 1:4
p <- x/sum(x)
temp <- rbind(x, p)
rownames(temp) <- c("X", "Prob")
temp
```

<u>answer</u>

```{r a5,comment=''}
sum(x*p)
```

## Question 6

A web site (www.medicine.ox.ac.uk/bandolier/band64/b64-7.html) for home pregnancy tests cites the following: When the
subjects using the test were women who collected and tested their own samples, the overall sensitivity was 75%.
Specificity was also low, in the range 52% to 75%. Assume the lower value for the specificity. Suppose a subject has a
positive test and that 30% of women taking pregnancy tests are actually pregnant. What number is closest to the
probability of pregnancy given the positive test?

<u>answer</u>

\[\begin{eqnarray*}\mathbb{P}\left(\textrm{Preg}\bigg| +\right) &=& 
\frac{\mathbb{P}\left(+\bigg|\textrm{Preg}\right)\cdot \mathbb{P}\left(\textrm{Preg}\right)}
{\mathbb{P}\left(+\bigg|\textrm{Preg}\right)\cdot \mathbb{P}\left(\textrm{Preg}\right) + 
\mathbb{P}\left(+\bigg|\textrm{Preg}^{c}\right)\cdot \mathbb{P}\left(\textrm{Preg}^{c}\right)}\\
&=& \frac{0.75\cdot0.3}{0.75\cdot0.3 + (1-0.52)\cdot(1-0.3)}\\&\approx&0.4\end{eqnarray*}\]