---
title: "Quiz 2"
author: "George Papadopoulos </br> pgeorgios8@gmail.com"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---
<style>body{text-align: justify}</style>

## Question 1

What is the variance of the distribution of the average an IID draw of $n$ observations from a population with mean
$\mu$ and variance $\sigma^2$.

<u>answer</u>

\[\frac{\sigma^2}{n}\]

## Question 2

Suppose that diastolic blood pressures (DBPs) for men aged 35-44 are normally distributed with a mean of 80 (mm Hg) and a
standard deviation of 10. About what is the probability that a random 35-44 year old has a DBP less than 70?

<u>answer</u>

```{r a2,comment=''}
pnorm(70,80,10)
```

## Question 3

Brain volume for adult women is normally distributed with a mean of about 1,100 cc for women with a standard deviation
of 75 cc. What brain volume represents the 95th percentile?

<u>answer</u>

```{r a3,comment=''}
qnorm(.95,1100,75)
```

## Question 4

Refer to the previous question. Brain volume for adult women is about 1,100 cc for women with a standard deviation of
75 cc. Consider the sample mean of 100 random adult women from this population. What is the 95th percentile of the
distribution of that sample mean? 

<u>answer</u>

```{r a4,comment=''}
qnorm(.95,1100,75/sqrt(100))
```

## Question 5

You flip a fair coin 5 times, about what's the probability of getting 4 or 5 heads?

<u>answer</u>

```{r a5,comment=''}
pbinom(3,5,0.5,lower.tail = FALSE)
```

## Question 6

The respiratory disturbance index (RDI), a measure of sleep disturbance, for a specific population has a mean of 15
(sleep events per hour) and a standard deviation of 10. They are not normally distributed. Give your best estimate of
the probability that a sample mean RDI of 100 people is between 14 and 16 events per hour?

<u>answer</u>

```{r a6,comment=''}
pnorm(16,15,10/sqrt(100)) - pnorm(14,15,10/sqrt(100))
```

## Question 7

Consider a standard uniform density. The mean for this density is .5 and the variance is 1 / 12. You sample 1,000
observations from this distribution and take the sample mean, what value would you expect it to be near?

<u>answer</u>

According to the law of large numbers (LLN) around 0.5.

## Question 8

The number of people showing up at a bus stop is assumed to be Poisson with a mean of 5 people per hour. You watch
the bus stop for 3 hours. About what's the probability of viewing 10 or fewer people?

<u>answer</u>

```{r a8,comment=''}
ppois(10,5*3)
```