---
title: "Quiz 4"
author: "George Papadopoulos </br> pgeorgios8@gmail.com"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---
<style>body{text-align: justify}</style>

## Question 1

A pharmaceutical company is interested in testing a potential blood pressure lowering medication.
Their first examination considers only subjects that received the medication at baseline then two weeks later.
The data are as follows (SBP in mm Hg)

|Subject|Baseline|Week 2|
|:------|:-------|:-----|
|1	    |140     |132   |
|2      |138     |135   |
|3	    |150	 |151   |
|4	    |148     |146   |
|5	    |135	 |130   |

Consider testing the hypothesis that there was a mean reduction in blood pressure?
Give the p-value for the associated two sided t-test.

<u>answer</u>

We will consider the observations for each subject as paired.

```{r a1,comment = ''}
bl <- c(140, 138, 150, 148, 135)
fu <- c(132, 135, 151, 146, 130)
t.test(fu,bl,alternative = "two.sided",paired = TRUE)$p.value
```

$\mathcal{H}_{0}:\mu=0$ vs $\mathcal{H}_{a}:\mu\neq0$, since the p-value is greater than 0.05 
we fail to reject the null hypothesis.

## Question 2

A sample of 9 men yielded a sample average brain volume of 1,100cc and a standard deviation of 30cc. What is the
complete set of values of $\mu_0$ that a test of $\mathcal{H}_{0}:\mu=\mu_{0}$ would fail to reject the null
hypothesis in a two sided 5% Students t-test?

<u>answer</u>

```{r a2,comment = ''}
1100 + c(-1,1) * qt(.975,8) * 30 / sqrt(9)
```

## Question 3

Researchers conducted a blind taste test of Coke versus Pepsi. Each of four people was asked which of two blinded
drinks given in random order that they preferred. The data was such that 3 of the 4 people chose Coke. Assuming that
this sample is representative, report a p-value for a test of the hypothesis that Coke is preferred to Pepsi using a
one sided exact test.

<u>answer</u>

```{r a3,comment = ''}
pbinom(2,size = 4,prob = 0.5,lower.tail = FALSE)
```

## Question 4

Infection rates at a hospital above 1 infection per 100 person days at risk are believed to be
too high and are used as a benchmark. A hospital that had previously been above the benchmark
recently had 10 infections over the last 1,787 person days at risk. About what is the one sided
p-value for the relevant test of whether the hospital is *below* the standard? 

<u>answer</u>

```{r a4,comment = ''}
ppois(10,lambda = 0.01 * 1787)
```

## Question 5

Suppose that 18 obese subjects were randomized, 9 each, to a new diet pill and a placebo. Subjects’
body mass indices (BMIs) were measured at a baseline and again after having received the treatment
or placebo for four weeks. The average difference from follow-up to the baseline (followup - baseline)
was −3 kg/m2 for the treated group and 1 kg/m2 for the placebo group. The corresponding standard
deviations of the differences was 1.5 kg/m2 for the treatment group and 1.8 kg/m2 for the placebo group.
Does the change in BMI appear to differ between the treated and placebo groups?  Assuming normality of 
the underlying data and a common population variance, give a p-value for a two sided t test.

<u>answer</u>

```{r a5,comment = ''}
n1 <- n2 <- 9
x1 <- -3 ##treated
x2 <- 1  ##placebo
s1 <- 1.5 ##treated
s2 <- 1.8 ##placebo
s <- sqrt(((n1 - 1) * s1^2 + (n2 - 1) * s2^2)/(n1 + n2 - 2))
ts <- (x1 - x2)/(s * sqrt(1/n1 + 1/n2))
2 * pt(ts,n1 + n2 - 2)
```

## Question 6

Brain volumes for 9 men yielded a 90% confidence interval of 1,077 cc to 1,123 cc.
Would you reject in a two sided 5% hypothesis test of $\mathcal{H}_{0}: \mu = 1,078$?

<u>answer</u>

No you wouldn't reject.

## Question 7

Researchers would like to conduct a study of 100 healthy adults to detect a four year mean brain
volume loss of $.01\;mm^3$. Assume that the standard  deviation of four year volume loss in this
population is  $.04\;mm^3$. About what would be the power of the study for a $5\%$ 
one sided test versus a null hypothesis of no volume loss?

<u>answer</u>

\[
\begin{eqnarray*}
\mathbb{P}\left(\frac{\overline{X_{\Delta}}}{\frac{\sigma_{\Delta}}{\sqrt{n}}}>\textbf{q}_{norm}(.95)\bigg|\mu_{\Delta} = 0.01\right) &=& 
\mathbb{P}\left(\frac{\overline{X_{\Delta}}-0.01}{\frac{\sigma_{\Delta}}{\sqrt{n}}}>1.645 -
\frac{0.01}{\frac{\sigma_{\Delta}}{\sqrt{n}}}\bigg|\mu_{\Delta} 
= 0.01\right)\\
&=&\mathbb{P}\left(\mathbb{Z} > 1.645 - \frac{.01}{\frac{.04}{\sqrt{100}}}\bigg|\mu_{\Delta} 
= 0.01\right) \\
&=& \mathbb{P}\left(\mathbb{Z}>-.855\right) \\
&\approx& .8 
\end{eqnarray*}
\]

```{r a7,comment = ''}
pnorm(1.645 * 0.004, mean = 0.01, sd = 0.004, lower.tail = FALSE)
```


## Question 8

Researchers would like to conduct a study of n healthy adults to detect a four year mean brain
volume loss of $.01 mm^3$. Assume that the standard deviation of four year volume loss in this population is 
$.04\;mm^3$. About what would be the value of n needed for $90\%$ power of type one error 
rate of $5\%$ one sided test versus a null hypothesis of no volume loss?

<u>answer</u>

\[
\begin{eqnarray*}
\mathbb{P}\left(\frac{\overline{X_{\Delta}}}{\frac{\sigma_{\Delta}}{\sqrt{n}}}>\textbf{q}_{norm}(.95)\bigg|\mu_{\Delta} = 0.01\right) &=& 0.9\\
\mathbb{P}\left(\frac{\overline{X_{\Delta}}-0.01}{\frac{\sigma_{\Delta}}{\sqrt{n}}}>1.645 -
\frac{0.01}{\frac{\sigma_{\Delta}}{\sqrt{n}}}\bigg|\mu_{\Delta} = 0.01\right) &=& 0.9\\
\mathbb{P}\left(\mathbb{Z} > 1.645 - \frac{\sqrt{n}}{4}\right) &=& 0.9 \\
1.645 - \frac{\sqrt{n}}{4} &=& \textbf{q}_{norm}(.1) \\
n &=& 16(\textbf{q}_{norm}(.1)+1.645)^2
\end{eqnarray*}
\]

```{r a8,comment = ''}
ceiling((4 * (qnorm(0.95) - qnorm(0.1)))^2)
```

## Question 9

As you increase the type one error rate, $\alpha$, what happens to power?

<u>answer</u>

You will get larger power.
