---
title: "Programming Assignment 1"
author: "George Papadopoulos </br> pgeorgios8@gmail.com"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---
<style>body{text-align: justify}</style>

## Introduction

For this first programming assignment you will write three functions that are meant to
interact with dataset that accompanies this assignment.

The zip file containing the data can be downloaded: 
[here]("https://d396qusza40orc.cloudfront.net/rprog%2Fdata%2Fspecdata.zip")

```{r data download unzip,echo = TRUE,eval = TRUE,cache = TRUE}
if (!file.exists("specdata")) {
  download.file("https://d396qusza40orc.cloudfront.net/rprog%2Fdata%2Fspecdata.zip","data.zip")
  unzip("data.zip")
}
```

The zip file contains 332 comma-separated-value (CSV) files containing pollution
monitoring data for fine particulate matter (PM) air pollution at 332 locations in the
United States. Each file contains data from a single monitor and the ID number for each
monitor is contained in the file name. For example, data for monitor 200 is contained in
the file "200.csv". Each file contains three variables:

* Date: the date of the observation in YYYY-MM-DD format (year-month-day)

* sulfate: the level of sulfate PM in the air on that date (measured in micrograms per
cubic meter)

* nitrate: the level of nitrate PM in the air on that date (measured in micrograms per
cubic meter)

For this programming assignment you will need to unzip this file and create the
directory 'specdata'. Once you have unzipped the zip file, do not make any modifications
to the files in the 'specdata' directory. In each file you'll notice that there are many
days where either sulfate or nitrate (or both) are missing (coded as NA). This is common
with air pollution monitoring data in the United States.

### Part 1

Write a function named 'pollutantmean' that calculates the mean of a pollutant (sulfate
or nitrate) across a specified list of monitors. The function 'pollutantmean' takes
three arguments: 'directory', 'pollutant', and 'id'. Given a vector monitor ID numbers,
'pollutantmean' reads that monitors' particulate matter data from the directory
specified in the 'directory' argument and returns the mean of the pollutant across all
of the monitors, ignoring any missing values coded as NA.

```{r function 1,echo = TRUE,cache = TRUE,eval = TRUE}
pollutantmean <- function(directory = "specdata", pollutant = "sulfate", id = 1:332) {
    ## 'directory' is a character vector of length 1 indicating
    ## the location of the CSV files
    
    ## 'pollutant' is a character vector of length 1 indicating
    ## the name of  the pollutant for which we will calcultate the
    ## mean; either "sulfate" or "nitrate"
    
    ## 'id' is an integer vector indicating the monitor ID numbers
    ## to be used
    
    ## Return the mean of the pollutant across all monitors list
    ## in the 'id' vector (ignoring NA values)
    ## NOTE: Do not round the result
  
  ## directory specified by the user
  ## pollutant must be either "sulfate" or "nitrate"
  ## id must be an integer between 1 and 332
  vec     <- c()
  for (x in id) {
    if ( x < 10) {
      name <- paste("00",x,sep = "")
    } else if ( x < 100) {
      name <- paste("0",x,sep = "")
    }
    else {
      name <- as.character(x)
    }
    read_file <- read.csv(paste(directory,"/",name,".csv",sep = ""))
    vec_helper      <- read_file[pollutant]
    vec_helper      <- vec_helper[!is.na(vec_helper)]
    vec <- c(vec,vec_helper)
  }
  return(mean(vec))
}
```

Testing:

```{r test 1,comment = ""}
pollutantmean("specdata", "sulfate", 1:10)
pollutantmean("specdata", "nitrate", 70:72)
pollutantmean("specdata", "nitrate", 23)
```

### Part 2

Write a function that reads a directory full of files and reports the number of completely observed cases in each
data file. The function should return a data frame where the first column is the name of the file and the second
column is the number of complete cases.

```{r function2,echo = TRUE,cache = TRUE,eval = TRUE}
complete <- function(directory = "specdata", id = 1:332) {
    ## 'directory' is a character vector of length 1 indicating
    ## the location of the CSV files
    
    ## 'id' is an integer vector indicating the monitor ID numbers
    ## to be used
    
    ## Return a data frame of the from:
    ## id nobs
    ## 1  117
    ## 2  1041
    ## ...
    ## where 'id' is the monitor ID number and 'nobs' is the
    ## number of complete cases
  
  ## directory specified by the user
  ## id must be an integer between 1 and 332
  
  data_frame <- as.data.frame(matrix(NA,length(id),2))
  count <- 1
  for (x in id) {
    # different way to get the names of the files
    file_name <- paste(directory,"/",sprintf("%03d",x),".csv",sep = "")
    cache     <- read.csv(file_name)
    data_frame[count,] <- c(x,sum(complete.cases(cache)))
    count <- count + 1
  }
  names(data_frame) <- c("id","nobs")
  return(data_frame)
}
```

Testing: 

```{r test 2,comment = ""}
complete("specdata", 1)
complete("specdata", c(2, 4, 8, 10, 12))
complete("specdata", 30:25)
complete("specdata", 3)
```

### Part 3

Write a function that takes a directory of data files and a threshold for complete cases and calculates the
correlation between sulfate and nitrate for monitor locations where the number of completely observed cases 
(on all variables) is greater than the threshold. The function should return a vector of correlations for the
monitors that meet the threshold requirement. If no monitors meet the threshold requirement, then the function
should return a numeric vector of length 0. For this function you will need to use the 'cor' function in R which
calculates the correlation between two vectors. Please read the help page for this function via '?cor' and make
sure that you know how to use it.


```{r function3,echo = TRUE,cache = TRUE,eval = TRUE}
corr <- function(directory = "specdata", threshold = 0) {
    ## 'directory' is a character vector of length 1 indicating
    ## the location of the CSV files
    
    ## 'threshold' is a numeric vector of length 1 indicating the 
    ## number of completely observed observations (on all
    ## variables) required to compute the correlation between
    ## nitrate and sulfate; the default is 0
    
    ## Return a numeric vector of correlations
    ## NOTE: Do not round the result!
  
  ## directory specified by the user
  vec <- c()
  for (x in 1:332) {
    ## alternative III: file_name <- dir(directory,full.names = TRUE)
    file_name <- paste(directory,"/",sprintf("%03d",x),".csv",sep = "")
    cache     <- read.csv(file_name)
    if (sum(complete.cases(cache)) >= threshold) {
      cache <- cache[complete.cases(cache),]
      vec <- c(vec,cor(cache[,2],cache[,3]))
    }
  }
  return(vec)
}
```


Testing:

```{r test3,comment = ""}
cr <- corr("specdata", 150)
head(cr)
summary(cr)
cr <- corr("specdata", 400)
head(cr)
summary(cr)
cr <- corr("specdata", 5000)
summary(cr)
length(cr)
cr <- corr("specdata")
summary(cr)
length(cr)
```