---
title: "Programming Assignment 3"
author: "George Papadopoulos <br /> pgeorgios8@gmail.com"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: leonids
    highlight: github
---
<style>body{text-align: justify}</style>

## Introduction

The  Hospital  Compare  web  site  contains  a  lot  of  data  and  we  will  only  look  at  a  small  subset
for this assignment.  The zip file for this assignment contains three files

* outcome-of-care-measures.csv: Contains information about 30-day mortality and readmission rates for heart 
attacks, heart failure, and pneumonia for over 4,000 hospitals.

* hospital-data.csv:  Contains information about each hospital.

* Hospital_Revised_Flatfiles.pdf:  Descriptions of the variables in each file (i.e the code book). A 
description of the variables in each of the files is in the included PDF file named 
Hospital_Revised_Flatfiles.pdf. This document contains information about many other files that are not included 
with this programming assignment.

You will want to focus on the variables for number 19 (“Outcome of Care Measures.csv”) and number 11 
(“HospitalData.csv”). You may find it useful to print out this document (at least the pages for Tables 19 and 
11) to have next to you while you work on this assignment. In particular, the numbers of the variables for each
table indicate column indices in each table (i.e. “Hospital Name” is column 2 in the outcome-of-care-measures.csv 
file).

### 1 Plot the 30-day mortality rates for heart attack

Read the outcome data into data into R via the read.csv function and look at the first few rows.

```{r head}
data <- read.csv("outcome-of-care-measures.csv",colClasses = "character")
#head(outcome)
```

There are many columns in this dataset.  You can see how many by typing ncol(outcome) (you can see the number of
rows with the __nrow__ function).  In addition, you can see the names of each column by typing names(outcome) (the
names are also in the PDF document.To make a simple histogram of the 30-day death rates from heart attack (column
11 in the outcome dataset).

```{r histogram}
data[,11] <- as.numeric(data[,11])
## you may get a warning about NAs being introduced; that is okay
hist(data[,11],xlab = "30 days mortality rates from heart attack",main = "",col = 'red')
```

Because we originally read the data in as character by specifying colClasses = "character" we need to coerce the 
column to be numeric. You may get a warning about NAs being introduced but that is okay.

```{r data preparation}
data <- data[,c(2,7,11,17,23)]
data[,4] <- as.numeric(data[,4])
data[,5] <- as.numeric(data[,5])
```
### 2 Finding the _best_ hospital in a state

Write a function called _best_ that take two arguments:  the 2-character abbreviated name of a _state_ and an
_outcome_ name. The function reads the outcome-of-care-measures.csv file and returns a character vector with  the 
name  of the hospital  that  has  the  _best_  (i.e.   lowest)  30-day  mortality  for  the  specified  _outcome_
in that _state_.  The hospital name is the name provided in the Hospital.Name variable.  The outcomes can be one of
“heart attack”, “heart failure”, or “pneumonia”.  Hospitals that do not have data on a particular _outcome_ should 
be excluded from the set of hospitals when deciding the rankings.

_Handling ties._ If there is a tie or the _best_ hospital for a given _outcome_, then the hospital names should be
sorted in alphabetical order and the first hospital in that set should be chosen (i.e.  if hospitals “b”, “c”,and
“f” are tied for _best_, then hospital “b” should be returned). The function should check the validity of its
arguments.  If an invalid _state_ value is passed to _best_,  the function should throw an error via the stop 
function with the exact message “invalid state”. If an invalid _outcome_ value is passed to _best_, the function 
should throw an error via the stop function with the exact message “invalid outcome”.

```{r function best}
best <- function(state,outcome) {
  if (!(state %in% unique(data$State))) {
    print("Valid state input:")
    cat(unique(data$State))
    cat('\n')
    stop("invalid state")
  }
  outcomes <- list('heart attack' = 3,'heart failure' = 4,'pneumonia' = 5)
  if (!(outcome %in% names(outcomes))) {
    print ("Valid outcomes:")
    print(names(outcomes))
    cat('\n')
    stop("invalid outcome")
  }
  data <- data[data$State == state,c(1,outcomes[[outcome]])]
  data <- data[!is.na(data[,2]),]
  low  <- which.min(data[,2])
  low  <- data[low,1]
  if(length(low) == 0) return(NA)
  return(sort(low))
}
```

Testing:

```{r part 1 test,comment = ""}
best("TX","heart attack")
best("TX","heart failure")
best("MD","heart attack")
best("MD","pneumonia")
best("SC","heart attack")
best("NY","pneumonia")
best("AK","pneumonia")
# best("NY","heart attack")  ---> throws error
# best("BB","heart attack") ---> throws error
```

### 3 Ranking hospitals by outcome in a state

Write a function called _rankhospital_ that takes three arguments: the 2-character abbreviated name of a
_state_, an _outcome_, and the ranking of a hospital in that state for that outcome (_num_).
The function reads the outcome-of-care-measures.csv and returns a character vector with the name
of the hospital that has the ranking specified by the _num_ argument. For example, the call

rankhospital("MD", "heart failure", 5)

would return a character vector containing the name of the hospital with the 5th lowest 30-day death rate
for heart failure. The _num_ argument can take values "_best_", "worst", or an integer indicating the ranking
(smaller numbers are better). If the number given by _num_ is larger than the number of hospitals in that
state, then the function should return $\color{red}{\text{NA}}$. Hospitals that do not have data on a particular
_outcome_ should be excluded from the set of hospitals when deciding the rankings.

Handling ties: It may occur that multiple hospitals have the same 30-day mortality rate for a given cause
of death. In those cases ties should be broken by using the hospital name. For example, in Texas ("TX"),
the hospitals with lowest 30-day mortality rate for heart failure are shown here.


|\>head(texas)||||
|:----|:-----:|:-----:|:--------:|
|Hospital.Name Rate Rank||||
|3935| FORT DUNCAN MEDICAL CENTER|8.1|1|
|4085| TOMBALL REGIONAL MEDICAL CENTER|8.5|2|
|4103| CYPRESS FAIRBANKS MEDICAL CENTER|8.7|3|
|3954| DETAR HOSPITAL NAVARRO|8.7|4|
|4010| METHODIST HOSPITAL,THE|8.8|5|
|3962| MISSION REGIONAL MEDICAL CENTER|8.8|6|

Note that Cypress Fairbanks Medical Center and Detar Hospital Navarro both have the same 30-day rate
(8.7). However, because Cypress comes before Detar alphabetically, Cypress is ranked number 3 in this
scheme and Detar is ranked number 4. One can use the order function to sort multiple vectors in this
manner (i.e. where one vector is used to break ties in another vector).

```{r function rankhospital}
rankhospital <- function(state,outcome,num = 'best') {
  if (!(state %in% unique(data$State))) {
    print("Valid state input:")
    cat(unique(data$State))
    cat('\n')
    stop("invalid state")
  }
  outcomes <- list('heart attack' = 3,'heart failure' = 4,'pneumonia' = 5)
  if (!(outcome %in% names(outcomes))) {
    print ("Valid outcomes:")
    print(names(outcomes))
    cat('\n')
    stop("invalid outcome")
  }
  if (is.character(num) & !(num == 'best' | num == 'worst')) {
    stop("num character input must be best or worst!")
  }
  if (is.numeric(num) & !((num) == num & num >= 1)) {
    stop("num numeric input must be an natural number!")
  }
  if (num == 'best') num <- 1
  data <- data[data$State == state,c(1,outcomes[[outcome]])]
  data <- data[!is.na(data[,2]),]
  if (num == 'worst') num <- dim(data)[1]
  names(data) <- c('hospital','rates')
  data <- dplyr::arrange(data,rates,hospital)
  return(data[num,1])
}
```

Testing:

```{r part 2 test,comment = ''}
rankhospital("TX", "heart failure", 4)
rankhospital("MD", "heart attack", "worst")
rankhospital("NC", "heart attack", "worst")
rankhospital("WA", "heart attack", 7)
rankhospital("TX", "pneumonia", 10)
rankhospital("NY", "heart attack", 7)
rankhospital("TX","pneumonia",10000)
```

### 4 Ranking hospitals in all states

Write a function called _rankall_ that takes two arguments: an _outcome_ name and a hospital ranking _num_. 
The function reads the outcome-of-care-measures.csv file and returns a 2-column data frame
containing the hospital in each state that has the ranking specified in _num_. For example the function call
_rankall_("heart attack", "best") would return a data frame containing the names of the hospitals that
are the _best_ in their respective states for 30-day heart attack death rates. The function should return a value
for every _state_ (some may be NA). The first column in the data frame is named hospital, which contains
the hospital name, and the second column is named _state_, which contains the 2-character abbreviation for
the _state_ name. Hospitals that do not have data on a particular _outcome_ should be excluded from the set of
hospitals when deciding the rankings.

Handling ties. The _rankall_ function should handle ties in the 30-day mortality rates in the same way
that the _rankhospital_ function handles ties.

__NOTE__: For the purpose of this part of the assignment (and for efficiency), your function should NOT call
the _rankhospital_ function from the previous section.
The function should check the validity of its arguments. If an invalid _outcome_ value is passed to _rankall_,
the function should throw an error via the stop function with the exact message "invalid outcome". The _num_
variable can take values "_best_", "worst", or an integer indicating the ranking (smaller numbers are better).
If the number given by num is larger than the number of hospitals in that _state_, then the function should
return __NA__.

```{r rankall}
rankall <- function(outcome,num = 'best') {
  outcomes <- list('heart attack' = 3,'heart failure' = 4,'pneumonia' = 5)
  if (!(outcome %in% names(outcomes))) {
    print ("Valid outcomes:")
    print(names(outcomes))
    cat('\n')
    stop("invalid outcome")
  }
  if (is.character(num) & !(num == 'best' | num == 'worst')) {
    stop("num character input must be best or worst!")
  }
  if (is.numeric(num) & !((num) == num & num >= 1)) {
    stop("num numeric input must be an natural number!")
  }
  states <- sort(unique(data$State))
  result <- data.frame(State = states,Hospital = NA)
  if (num == 'best')  num <- 1
  count <- 1
  for (state in states) {
    cache <- data[data$State == state,c(1,2,outcomes[[outcome]])]
    cache <- cache[!is.na(cache[,3]),]
    cache <- cache[,c(1,3)]
    names(cache) <- c('hospital','rates')
    cache <- dplyr::arrange(cache,rates,hospital)
    if (num == 'worst') {
      result[count,2] <- cache[dim(cache)[1],1]
    } else {
      result[count,2] <- cache[num,1]
    }
    count <- count + 1
  }
  return(result)
}
```

```{r part 3,comment = ''}
head(rankall("heart attack", 20), 10)
tail(rankall("pneumonia", "worst"), 3)
tail(rankall("heart failure"), 10)
r <- rankall("heart attack", 4)
as.character(subset(r, State == "HI")$Hospital)
r <- rankall("pneumonia", "worst")
as.character(subset(r, State == "NJ")$Hospital)
r <- rankall("heart failure", 10)
as.character(subset(r, State == "NV")$Hospital)
```