---
title: "Q2"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE,comment = '')
```
<style>
  body {
    text-align: justify
    }
</style>

### 1.

<b>Suppose I define the following function in R,</b>

```{r w1,eval = FALSE,echo = TRUE}
cube <- function(x, n) {
        x^3
}
```

<b>What is the result of running</b>

```{r w2,eval = FALSE,echo = TRUE}
cube(3)
```

<b>in R after defining this function?</b>


```{r 1}
question('',
  answer('27',correct = TRUE),
  answer("An error is returned because 'n' is not specified in the call to 'cube'"),
  answer("The users is prompted to specify the value of 'n'."),
  answer("A warning is given with no value returned."))
```

### 2.

<b>The following code will produce a warning in R. Why?</b>

```{r w3,eval = FALSE,echo = TRUE}
x <- 1:10
if(x > 5) {
        x <- 0
}
```

```{r q2}
question('',
  answer("There are no elements in 'x' that are greater than 5"),
  answer("The expression uses curly braces."),
  answer("You cannot set 'x' to be 0 because 'x' is a vector and 0 is a scalar."),
  answer("The syntax of this R expression is incorrect."),
  answer("'x' is a vector of length 10 and 'if' can only test a single logical statement.",
         correct = TRUE))
```

### 3.

<b>Consider the following code.</b>

```{r w4,eval = FALSE,echo = TRUE}
f <- function(x) {
        g <- function(y) {
                y + z
        }
        z <- 4
        x + g(x)
}
```

<b>If then I run in R</b>

```{r w5,eval = FALSE,echo = TRUE}
z <- 10
f(3)
```

<b>what value is returned?</b>

```{r q3}
question('',
  answer('4'),
  answer('16'),
  answer('10',correct = TRUE),
  answer('7'))
```

### 4.

<b>Consider the following expression.</b>

```{r w6,eval = FALSE,echo = TRUE}
x <- 5
y <- if(x < 3) {
        NA
} else {
        10
}
```

<b>What is the value of 'y' after evaluating this expression?</b>

```{r q4}
question('',
  answer('3'),
  answer('10',correct = TRUE),
  answer('5'),
  answer('NA'))
```

### 5.

<b>Consider the following R function,</b>

```{r w7,eval = FALSE,echo = TRUE}
h <- function(x, y = NULL, d = 3L) {
        z <- cbind(x, d)
        if(!is.null(y))
                z <- z + y
        else
                z <- z + f
        g <- x + y / z
        if(d == 3L)
                return(g)
        g <- g + 10
        g
}
```

<b>which symbol in the above function is a free variable?</b>

```{r q5}
question('',
  answer('f',correct = TRUE),
  answer('z'),
  answer('d'),
  answer('L'),
  answer('g'))
```

### 6.

```{r q6}
question('What is an environment in R?',
  answer('a collection of symbol/value pairs',correct = TRUE),
  answer('an R package that only contains data'),
  answer('a list whose elements are all functions'),
  answer('a special type of function'))
```

### 7.

```{r q7}
question('The R language uses what type of scoping rule for resolving free variables?',
  answer('lexical scoping',correct = TRUE),
  answer('dynamic scoping'),
  answer('compilation scoping'),
  answer('global scoping'))
```

### 8.

```{r q8}
question('How are free variables in R functions resolved?',
  answer('The values of free variables are searched for in the environment in which the function was called'),
  answer('The values of free variables are searched for in the environment in which the function was defined',correct = TRUE),
  answer('The values of free variables are searched for in the global environment'),
  answer('The values of free variables are searched for in the working directory'))
```

### 9.

```{r q9}
question('What is one of the consequences of the scoping rules used in R?',
  answer('All objects must be stored in memory',correct = TRUE),
  answer('All objects can be stored on the disk'),
  answer('Functions cannot be nested'),
  answer('R objects cannot be larger than 100 MB'))
```

### 10.

```{r q10}
question('In R what is the parent frame?',
  answer('It is the package search list'),
  answer('It is the environment in which a function was defined'),
  answer('It is always the global environment'),
  answer('It is the environment in which a function was called',correct = TRUE))
```