---
title: "Programming Assignment 2"
author: "George Papadopoulos </br> pgeorgios8@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document: 
    css: style.css
    latex_engine: xelatex
    highlight: haddock
    theme: united
    toc: yes
---

## Introduction to the assignment

In this example we introduce the `<<-` operator which can be used to assign a value to an object in an environment
that is different from the current environment. Below are two functions that are used to create a special object
that stores a numeric vector and caches its mean.

The first function, __makeVector__ creates a special "vector", which is really a list containing a function to

* set the value of the vector </br>
* get the value of the vector </br>
* set the value of the mean   </br>
* get the value of the mean   </br>

```{r schema 1,eval = FALSE}
makeVector <- function(x = numeric()) {
        m <- NULL
        set <- function(y) {
                x <<- y
                m <<- NULL
        }
        get <- function() x
        setmean <- function(mean) m <<- mean
        getmean <- function() m
        list(set = set, get = get,
             setmean = setmean,
             getmean = getmean)
}
```

The following function calculates the mean of the special "vector" created with the above function. However, it
first checks to see if the mean has already been calculated. 
If so, it gets the mean from the cache and skips the
computation. Otherwise, it calculates the mean of the data and sets the value of the mean in the cache via the
__setmean__ function.


```{r schema 2,eval = FALSE}
cachemean <- function(x,...) {
        m <- x$getmean()
        if(!is.null(m)) {
                message("getting cached data")
                return(m)
        }
        data <- x$get()
        m <- mean(data, ...)
        x$setmean(m)
        m
}
```

## Assignment: Caching the Inverse of a Matrix

Matrix inversion is usually a costly computation and there may be some benefit to
caching the inverse of a matrix
rather than compute it repeatedly (there are also alternatives to matrix inversion 
that we will not discuss here).
Your assignment is to write a pair of functions that cache the inverse of a matrix.

Write the following functions:

* __makeCacheMatrix__: This function creates a special "matrix" object that can cache its inverse. </br>
* __cacheSolve__: This function computes the inverse of the special "matrix" returned by __makeCacheMatrix__ above. 
If the inverse has already been calculated (and the matrix has not changed),
then the __cachesolve__ should retrieve the inverse from the cache.</br>
* Computing the inverse of a square matrix can be done with the solve function in R.
For example, if X is a square invertible matrix, then __solve__(X) returns its inverse.

```{r makeCacheMatrix/cacheSolve}
makeCacheMatrix <- function(x = matrix()) {
  inv <- NULL
  get <- function() x
  set <- function(y) {
    x <<- y
    inv <<- NULL
  }
  getinv <- function() inv
  setinv <- function(inverse) {
    inv <<- inverse
  }
  return(list(
    set = set,
    get = get,
    getinverse = getinv,
    setinverse = setinv
  ))
}
cacheSolve <- function(x,...) {
  inverse <- x$getinverse()
  if (!is.null(inverse)) {
    return(inverse)
  }
  m <- solve(x$get())
  x$setinverse(m)
  #return(m)
}
```


<u>Explanation</u>:

* Constructor function "__makeCacheMatrix__"

This is a function that caches the computed values or the inputs. 
In the second line we create the variable _inv_ and set it by default to __NULL__. 


```{r diagram,fig.align = "center"}
library(DiagrammeR)
nodes <- create_node_df(
  n = 10,
  type  = c(rep("m",5),rep('o',3),rep("c",2)),
  label = c("set","get","get inverse","set inverse","cache constructor",
            "object in RAM","matrix","inverse matrix",
            "solve function","if not cached solve/save"),
  color = c(rep('red',5),rep('blue',3),rep('green',2))
)
edges <- create_edge_df(
  c(1,2,3,4,5,6,7,8,6,9,10),
  c(5,5,5,5,6,5,6,6,9,6,9),
  color = c(rep('red',5),rep('blue',4),rep('green',2)))
graph <- create_graph(
    nodes_df = nodes,
    edges_df = edges,attr_theme = "lr",directed = FALSE)

# View the graph
render_graph(graph,title = "Mind Map")
```

The first time that we run the function __makeCacheMatrix__ we create an object with 4 objects attached to it

* the matrix value which can be accessed with the get function of the same object

* set function that can change the cached value of matrix

* get inverse function to get the cached inverse of the matrix, 
which sets its value to NULL every time we assign a new matrix

* set inverse to get access to the object by the __cacheSolve__ function

* the `<<-` operator in both set functions attached to the object, is used to assign 
values in the parent frame outside of the function's environment created by __makeCacheMatrix__ function.

the __cacheSolve__ function on the other hand gets access to the object's inverse
matrix with the `<-` operator which "searches in the function's environment and then 
searches one level up in the parent environment"
and returns the cached inverse matrix without computing it again or computes the inverse otherwise for a
first entry matrix or if the matrix was changed with the __makeCacheMatrix$set(.)__ function.

```{r test,comment = ''}
I <- rbind(c(0,1,2),c(1,1,1),c(-4,2,1))
m <- makeCacheMatrix(I)
identical(m$get(),I)
## the inverse has not been computed yet
m$getinverse()
cacheSolve(m)
## now cacheSolve found out that the inverse in the object is NULL (<<- going one level up) 
## and replace the NULL value with the inverse of the matrix
m$getinverse()
## now let's set another matrix in the m object
m$set(rbind(sqrt(c(2,2)),c(-1,2)))
## notice that every time we enter a new matrix the inverse is turned to NULL with the <<- operator that 
## changes the m object's inverse value in the parent frame
m$getinverse()
```