---
title: "Clustering Example Motion Detection"
author: "George Papadopoulos </br> pgeorgios8@gmail.com"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: darkly 
    highlight: espresso
    toc: true
abstract: >
  "Recognizing human activity is a technological breakthrough of machine learning 
   paving the way to understanding and making the quality of the lives of people better. This 
   brief analysis is based on Sean Kross's swirl course _Exploratory Data Analysis_."
---

<style>
  body {
    text-align: justify
    }
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = '')
set.seed(1)
```

# Exploratory Data Analysis For Machine Learning

##  A Few Words/Introduction

The main goal of this brief report is to highlight the fruitful directions of research towards
answerable questions. An aspect of research which will be made clear
is that the "real-world" research isn't always neat and well-defined like textbook questions with 
theoretical clear cut questions. The question we will try to answer before training a machine learning 
model is "Is the correlation between the measurements and activities good enough to train a
machine?" so that "Given a set of 561 measurements, would a trained machine be able to determine which 
of the 6 activities the person was doing?"

## Data

The data we will use is already in a tidy format and contains 30 volunteers who performed 
activities of daily living while carrying a waist-mounted smartphone with embedded inertial sensors. 
Each person performed six activities while wearing a _Galaxy S II_ which were video recorded and so
during the performance of each activity the manually written  labels address the activity of the 
subject correctly. 

## Descriptive Analysis {.tabset .tabset-pills}

### Load Data

```{r load data}
ssd <- readRDS("samsung.rda")
```

### Specifications 

The data table  has `r dim(ssd)[1]` rows and `r dim(ssd)[2]` columns, meaning that we have 
7352 observations and 561 features since the last two columns store the subject id and the activity 
of the corresponding subject.

### Exploration With R

```{r names/dim}
dim(ssd)
names(ssd)
```

### Tables Activity/Subjects

```{r tables}
table(ssd$subject)
table(ssd$activity)
```

### Table Summation 

Sum the tables of observations per subject and activity to verify that they sum up to the number of
rows.

```{r sum/rows}
sum(table(ssd$subject))
sum(table(ssd$activity))
```

## Exploratory Analylis Part I

### Feature Plot By Index

For the purpose of this section we will subset the train dataset to limit it to only subject 1,

```{r subset}
sub1 <- subset(ssd,subject == 1)
dim(sub1)
```

we observe that the dataset with which we will work our analysis has only 347 observations. Let's 
review the names of the first 12 features again

```{r 12 first cols sub1}
names(sub1[1:12])
```

and focus on the first three which measure the mean of the acceleration in the three dimensions.

```{r plot 1,fig.align = "center"}
par(mfrow=c(2,3), mar = c(5, 4, 1, 1),oma = c(1,1,3,1))
plot(sub1[, 1], col = sub1$activity, ylab = names(sub1)[1])
plot(sub1[, 2], col = sub1$activity, ylab = names(sub1)[2])
plot(sub1[, 3], col = sub1$activity, ylab = names(sub1)[3])
plot(sub1[,1],sub1[,2],col = sub1$activity,xlab = names(sub1)[1],ylab = names(sub1)[2])
plot(sub1[,1],sub1[,3],col = sub1$activity,xlab = names(sub1)[1],ylab = names(sub1)[3])
plot(sub1[,2],sub1[,3],col = sub1$activity,xlab = names(sub1)[2],ylab = names(sub1)[3])
legend("bottomleft",legend=unique(sub1$activity),col=unique(sub1$activity), pch = 19)
mtext("Mean Acceleration Plots By Index And Against Each Other",outer = TRUE)
par(mfrow=c(1,1))
```

We observe that the active activities show more variability than the passive ones, especially 
in the X dimension. 

### Clustering 

We still focus on the three dimensions of the _mean acceleration_ features.

```{r clust 1}
mdist <- dist(sub1[,1:3])
hclustering <- hclust(mdist)
```

We will define a function for a more beautiful hierarchical clustering plot as a 
dendrogram written by Eva KF Chan.

```{r myplclust}
myplclust <- function(hclust,lab = hclust$labels,lab.col = rep(1,length(hclust$labels)), 
                       hang = 0.1,...){
 ## modification of plclust for plotting hclust objects *in colour*!
 ## Copyright Eva KF Chan 2009
 ## Arguments:
 ##    hclust:     hclust object
 ##    lab:        a character vector of labels of the leaves of the tree
 ##    lab.col:    color for the labels; NA=default device foreground color
 ##    hang:       as in hclust & plclust
 ## Side effect:
 ##    A display of hierarchical cluster with colored leaf labels.
 y <- rep(hclust$height,2)
 x <- as.numeric(hclust$merge)
 y <- y[which(x < 0)]
 x <- x[which(x < 0)]
 x <- abs(x)
 y <- y[order(x)]
 x <- x[order(x)]
 plot(hclust,labels = FALSE,hang = hang,...)
 text(x = x,y = y[hclust$order] - (max(hclust$height) * hang),labels = lab[hclust$order],
       col = lab.col[hclust$order],srt = 90,adj = c(1,0.5),xpd = NA,...)}
```

We will plot the hierarchical clustering as a dendrogram, based on the distance matrix we computed,

```{r cluster plot1,fig.align = "center"}
myplclust(hclustering,lab.col = unclass(sub1$activity),xlab = "activity")
legend("topright",legend=unique(sub1$activity),col=unique(sub1$activity),pch = 19)
```

we observe that we cannot understand much from the average acceleration since all colors seem 
jumbled together and actually there is no sign of clustering at all, so we need to explore another 
triplet of features which might be more useful.

## Exploratory Analysis Part II

### Feature Plot By Index

Since we didn't figure out much from the first three features, let's pick the columns 10 through 12 
which measure the maximum acceleration of a subject in three dimensions. As in part I let's plot the 
maximum acceleration as stored in 10th, 11th and 12th column.

```{r plot 2,fig.align = "center"}
par(mfrow=c(2, 3), mar = c(5, 4, 1, 1),oma = c(1,1,3,1))
plot(sub1[, 10], col = sub1$activity, ylab = names(sub1)[10])
plot(sub1[, 11], col = sub1$activity, ylab = names(sub1)[11])
plot(sub1[, 12], col = sub1$activity, ylab = names(sub1)[12])
plot(sub1[,10],sub1[,11],col = sub1$activity,xlab = names(sub1)[10],ylab = names(sub1)[11])
plot(sub1[,10],sub1[,12],col = sub1$activity,xlab = names(sub1)[10],ylab = names(sub1)[12])
plot(sub1[,11],sub1[,12],col = sub1$activity,xlab = names(sub1)[11],ylab = names(sub1)[12])
legend("topleft",legend=unique(sub1$activity),col=unique(sub1$activity),pch = 19)
mtext("Max Acceleration Plots By Index And Against Each Other",outer = TRUE)
par(mfrow=c(1,1))
```

We're seeing something vaguely interesting, passive activities mostly fall below the active ones!

### Hierarchical Clustering

As in part I we will create a _distance_ matrix

```{r clust 2}
mdist <- dist(sub1[,10:12])
hclustering <- hclust(mdist)
```

and create the hierarchical clustering plot

```{r cluster plot2,fig.align = "center"}
myplclust(hclustering,lab.col = unclass(sub1$activity),xlab = "activity")
legend("topright",legend=unique(sub1$activity),col=unique(sub1$activity),pch = 19)
```

we see clearly that the data splits into 2 clusters, active and passive activities. Moreover, the
light blue (walking down) is clearly distinct from the other walking activities. The dark blue (walking
level) also seems to be somewhat clustered. The passive activities, however, seem all jumbled together
with no clear pattern visible.

### Singular Value Decomposition

We will try __SVD__ on the dataset sub1, the argument we will pass will be a scaled data frame with 
the last two columns removed (subject id and activity).

```{r svd}
svd1 <- svd(scale(sub1[,-c(562,563)]))
```

The first two principal components explain 64% of the data.

```{r barlot,fig.align = 'center'}
barplot(svd1$d[1:10]^2/sum(svd1$d^2),names = mapply("paste0","PC",1:10),las = 3,col = 'red')
```

We will now plot the first two left singular vectors of svd1 which are  the first two columns of the 
U matrix, remember that __SVD__ transformed matrix sub1 into $UDV^{T}$ and that each row in U 
corresponds to a row in the matrix sub1.

```{r plot two singular U,fig.align = "center"}
par(mfrow = c(1,2),mar = c(1,1,1,6))
plot(svd1$u[,1],pch = 19,col = unclass(sub1$activity))
legend("right",inset = c(-.45,0),legend=unique(sub1$activity),
       col=unique(sub1$activity),pch = 19,xpd = TRUE)
plot(svd1$u[,2],pch = 19,col = unclass(sub1$activity))
```

Here we're looking at the 2 left singular vectors of svd1 (the first 2 columns of $U$). Each entry of
the columns belongs to a particular row with one of the 6 activities assigned to it. We see the
activities distinguished by color. Moving from left to right, the first section of rows are green
(standing), the second red (sitting), the third black (laying), etc.  The first column of U shows
separation of the non moving (black, red, and green) from the walking activities. The second column is
harder to interpret. However, the magenta cluster, which represents __walking up__, 
seems separate from the others.

We'll try to figure out why that is. To do that we'll have to find which of the 500+ measurements
(represented by the columns of sub1) contributes to the variation of that component. Since we're
interested in sub1 columns, we'll look at the RIGHT singular vectors (the columns of V), and in
particular, the second one since the separation of the magenta cluster stood out in the second column of
$U$. We will plot the second principal component of V,

```{r plot v 2nd,fig.align = "center"}
plot(svd1$v[,2],col = rgb(0,0,.5,.4),pch = 19)
```

from the plot we don't see any pattern or anything else useful. In order to investigate the magenta 
colored cluster we will find the feature which contributes most to this separation by finding the 
maximum index of the second column of V.

```{r maxCon}
maxCon <- which.max(svd1$v[,2])
```

We will now create a distance matrix of the sub1 matrix which contains the 10th to 12th column and 
maxCon.

```{r clust 3}
mdist <- dist(sub1[,c(10:12,maxCon)])
```

Next we will plot the hierarchical clustering dendrogram,

```{r cluster plot 3,fig.align = "center"}
hclustering <- hclust(mdist)
myplclust(hclustering,lab.col = unclass(sub1$activity))
legend("topright",legend=unique(sub1$activity),col=unique(sub1$activity),pch = 19)
```

and we also need to find name of this magenta clustering contributor.

```{r which max name}
names(sub1[maxCon])
```

So the mean body acceleration in the frequency domain in the Z direction is the main contributor to this
clustering phenomenon we're seeing. Let's move on to k-means clustering to see if this technique can
distinguish between the activities.

### K-means Clustering

Since the labeled activities are six we will use the _kmeans_ function with input the 
sub1 matrix having removed the last two columns.

```{r kmeans sub1}
kClust <- kmeans(sub1[,-c(562,563)],centers = 6,nstart = 100)
```

Let's take a look at a how the activities are clustered around each center,

```{r kmeans table,fig.align = "center",fig.cap = "cluster table"}
table(kClust$cluster,sub1$activity)
```

We see that with 100 random starts, the passive activities tend to cluster together. One of the
clusters contains only laying, but in another cluster, standing and sitting group together. We will 
also check the dimensions of the centers,

```{r dim center}
dim(kClust$centers)
```

we see that the centers are a 6 by 561 array. Sometimes it is a good idea to look at the 
features of these centers to see if any dominate. We will do this for the laying activity which
as seen in the above has a size of 29.

```{r laying}
laying <- which(kClust$size == 29)
```

We will plot the first 12 features of the laying center,

```{r plot laying center,fig.align = "center"}
plot(kClust$centers[laying,1:12],pch = 19,ylab = "Laying Center",col = "orange")
```

from the plot we see the first three columns __dominate__ the laying cluster and we also need to 
find the names of these features

```{r laying dominating features}
names(sub1[1:3])
```

so the three direction of mean body acceleration seem to have the biggest effect on lying. We will 
do the same for the walkdown activity, 

```{r walkdown center cluster index}
walkdown <- which(kClust$size == 49)
```

we will now plot the first twelve columns of that center

```{r plot walkdonw center,fig.align = "center"}
plot(kClust$centers[walkdown,1:12],ylab = "Walkdown Center",pch = 19,col = "orange")
```

we see an interesting pattern, from left to right we are looking at the twelve accelaration 
measurements in groups of three in which the points decrease in value. The X direction 
dominates, then the Y and finally Z in this pattern on and on.