---
title: "Quiz 2"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)
```
<style>
  body {
    text-align: justify
    }
</style>

### 1.

<b>Load the Alzheimer's disease data using the commands:</b>

```{r echo=TRUE,comment = ''}
library(AppliedPredictiveModeling)
data(AlzheimerDisease)
```

```{r q1}
question("Which of the following commands will create non-overlapping training and test sets with about
         50% of the observations assigned to each?",
         answer("adData = data.frame(diagnosis,predictors)</br>
         testIndex = createDataPartition(diagnosis, p = 0.50,list=FALSE)</br>
         training = adData[-testIndex,]</br>
         testing = adData[testIndex,]",correct = TRUE),
         answer("adData = data.frame(diagnosis,predictors)</br>
         train = createDataPartition(diagnosis, p = 0.50,list=FALSE)</br>
         test = createDataPartition(diagnosis, p = 0.50,list=FALSE)"),
         answer("adData = data.frame(diagnosis,predictors)</br>
         trainIndex = createDataPartition(diagnosis,p=0.5,list=FALSE)</br>
         training = adData[-trainIndex,]</br>
         testing = adData[-trainIndex,]"),
         answer("adData = data.frame(diagnosis,predictors)</br>
         trainIndex = createDataPartition(diagnosis,p=0.5,list=FALSE)</br>
         training = adData[trainIndex,]</br>
         testing = adData[trainIndex,]"))
```

### 2.

<b>Load the cement data using the commands:</b>

```{r,echo=TRUE,comment = ''}
library(AppliedPredictiveModeling)
data(concrete)
library(caret)
set.seed(1000)
inTrain  <- createDataPartition(mixtures$CompressiveStrength, p = 3/4)[[1]]
training <- mixtures[ inTrain,]
testing  <- mixtures[-inTrain,]
```

Make a plot of the outcome (CompressiveStrength) versus the index of the samples. Color by each of the
variables in the data set (you may find the cut2() function in the Hmisc package useful for turning
continuous covariates into factors). 

```{r plot2,echo = TRUE,comment = ''}
## Since there are three types of cement based on compressive strength 
## we will color based on a cut of the compressiveStrength in breaks of three
## and base the color in each case on this cut
par(mfrow = c(1,3))
cols <- colorRampPalette(c('red','yellow'))
for (ii in names(training)[1:3]) {
  plot(training$CompressiveStrength,col = cols(3)[cut(training[,ii],breaks = 3)],
       main = paste('color by',ii),pch = 19,ylab = 'comp. strength')
}
for (ii in names(training)[4:6]) {
  plot(training$CompressiveStrength,col = cols(3)[cut(training[,ii],breaks = 3)],
       main = paste('color by',ii),pch = 19,ylab = 'comp. strength')
}
for (ii in names(training)[7:9]) {
  plot(training$CompressiveStrength,col = cols(3)[cut(training[,ii],breaks = 3)],
       main = paste('color by',ii),pch = 19,ylab = 'comp. strength')
}
par(mfrow = c(1,1))
```

```{r q2}
question('What do you notice in these plots?',
         answer("There is a non-random pattern in the plot of the outcome versus index that is perfectly
         explained by the FlyAsh variable so there may be a variable missing."),
         answer("There is a non-random pattern in the plot of the outcome versus index."),
         answer("There is a non-random pattern in the plot of the outcome versus index that is perfectly
         explained by the Age variable so there may be a variable missing."),
         answer("There is a non-random pattern in the plot of the outcome versus index that does not
         appear to be perfectly explained by any predictor suggesting a variable may be missing.",
         correct = TRUE))
```

### 3.

<b>Using the data loaded in Q2, make a histogram and confirm the SuperPlasticizer variable is
skewed. Normally you might use the log transform to try to make the data more symmetric. Why would that 
be a poor choice for this variable?</b>

```{r}
hist(training$Superplasticizer,col = 'red',ylab = 'superplasticizer',main = '',xlab = '')
```

```{r q3}
question("",
        answer("The SuperPlasticizer data include negative values so the log transform can not be
               performed."),
        answer("There are values of zero so when you take the log() transform those values will be
               -Inf.",correct = TRUE),
        answer("The log transform produces negative values which can not be used by some classifiers."),
        answer("The log transform does not reduce the skewness of the non-zero values of
               SuperPlasticizer"))
```

### 4.

<b>We have loaded the Alzheimer's disease data in Q1. Using the commands:</b>

```{r,echo = TRUE,comment = ''}
set.seed(3433)
adData  <- data.frame(diagnosis,predictors)
inTrain <- createDataPartition(adData$diagnosis, p = 3/4)[[1]]
training <- adData[ inTrain,]
testing  <- adData[-inTrain,]
```

<b>Find all the predictor variables in the training set that begin with IL. Perform principal components
on these variables with the preProcess() function from the caret package. Calculate the number of
principal components needed to capture 90% of the variance.</b>

```{r,echo = TRUE,comment = ''}
IL.predictors <- grep("^IL",names(training),ignore.case = TRUE,value = TRUE)
training <- training[,IL.predictors]
preProcess(training,method = 'pca',thresh = 0.9)
```

```{r q4}
question("Find all the predictor variables in the training set that begin with IL. Perform principal
         components on these variables with the preProcess() function from the caret package. Calculate
         the number of principal components needed to capture 90% of the variance. 
         How many are there?",
         answer("5"),
         answer("12"),
         answer("10"),
         answer("9",correct = TRUE))
```

### 5.

<b>With the Alzheimer's data again, using the commands:</b>

```{r,echo = TRUE,comment = ''}
## This doesn't work with the lastest R so we will use an older rng 
## RNGversion("3.0.0")
library(caret)
library(AppliedPredictiveModeling)
set.seed(3433)
data(AlzheimerDisease)
adData = data.frame(diagnosis,predictors)
inTrain = createDataPartition(adData$diagnosis, p = 3/4)[[1]]
training = adData[ inTrain,]
testing = adData[-inTrain,]
```

<b>Create a training data set consisting of only the predictors with variable names beginning with IL
and the diagnosis. Build two predictive models, one using the predictors as they are and one using PCA
with principal components explaining 80% of the variance in the predictors. Use method="glm" in the
train function.</b>

```{r echo = TRUE,comment = ''}
training.new <- training[,grep("^(IL|diagnosis)",names(training),value = TRUE,ignore.case = TRUE)] 
testing.new  <- testing[,grep("(IL|diagnosis)",names(testing))]
nopca_preproc  <- train(diagnosis ~ .,data = training.new,method = 'glm')
confusionMatrix(predict(nopca_preproc,testing.new),testing$diagnosis)
training.pca <- preProcess(training.new,method = 'pca',thres = 0.8)
training.pca <- predict(training.pca,training.new)
testing.pca  <- preProcess(testing.new,method = 'pca',thres = 0.8)
testing.pca  <- predict(testing.pca,testing.new)
pca_preproc  <- train(diagnosis ~ .,data = training.pca,method = 'glm')
confusionMatrix(predict(pca_preproc,testing.pca),testing$diagnosis)
```

```{r q5}
question("What is the accuracy of each method in the test set? Which is more accurate?",
         answer("Non-PCA Accuracy: 0.65</br>
         PCA Accuracy: 0.72",correct = TRUE),
         answer("Non-PCA Accuracy: 0.75</br>
         PCA Accuracy: 0.71"),
         answer("Non-PCA Accuracy: 0.72</br>
         PCA Accuracy: 0.65"),
         answer("Non-PCA Accuracy: 0.72</br>
         PCA Accuracy: 0.71"))
```